Forecasting the 2024 U.S. Presidential Election: A Poll-Based Approach

### 2024-11-02 paper/paper.qmd
Clarified methodology

@@

In this study, we employ two main models to forecast the outcome of the 2024 U.S. Presidential Election using polling data for the candidates Kamala Harris and Donald Trump. The models, a Baseline Model (Winner-Take-All National Model) and a Primary Model (Swing State-Based Model), provide two distinct but complementary approaches to assessing candidate support. The Baseline Model aggregates national polling data to project which candidate is likely to win based on national polling trends and assigns all electoral votes to the candidate with the higher projected national percentage on election day. This approach offers a straightforward prediction of overall voter sentiment across the country and captures the aggregate trend of support for each candidate as the election approaches.

On the other hand, the Primary Model is designed to provide a more granular projection by focusing on polling data at the state level, particularly for key swing states. This model assumes that the winner in each swing state can be determined by state-specific polling trends, and it projects support for each candidate within these competitive regions. By combining outcomes from both swing states and solid states, this model allows for a more nuanced prediction that aligns with the actual mechanics of the Electoral College. Together, these models are well-suited to address both the general national polling trends and the crucial role of swing states in determining the final electoral outcome.

@@

  if (nrow(candidate_data) > 0) {  # Check if there are rows for the candidate
    model <- lm(pct ~ days_towards_election, data = candidate_data)
    tidy_model <- tidy(model)  # Get a tidy summary of the model
    tidy_model$candidate <- candidate  # Add candidate to the output
    return(tidy_model)
  } else {
    return(NULL)  # Return NULL if no data for the candidate
  }
}

@@

# Combine results for all candidates into a single data frame
final_results_baseline <- bind_rows(results, .id = "candidate")

# View the final results
kable(final_results_baseline)

@@

## 1. Baseline Model (Winner-Take-All National Model)
The baseline model aggregates national polling data to predict a winner, with all electoral votes awarded to the candidate with the higher national polling percentage.

1. **National Polling Projections**  
   Let:
   - $P_H(t)$: Harris’s national polling percentage at $t$ days towards the election.
   - $P_T(t)$: Trump’s national polling percentage at $t$ days towards the election.

@@

2. **Election Day Prediction**  
   On election day:
   $$
   P_H(0) = \alpha_H, \quad P_T(0) = \alpha_T
   $$

3. **Winner-Take-All Outcome**  
   The candidate with the higher predicted national percentage wins all electoral votes:
   $$
   W = 
   \begin{cases} 
   \text{Harris wins (all electoral votes)} & \text{if } \alpha_H > \alpha_T \\
   \text{Trump wins (all electoral votes)} & \text{otherwise}
   \end{cases}
   $$

@@

  if (nrow(candidate_data) > 0) {  # Check if there are rows for the candidate
    model <- lm(pct ~ days_towards_election, data = candidate_data)
    tidy_model <- tidy(model)  # Get a tidy summary of the model
    tidy_model$state <- unique(candidate_data$state)  # Add state to the output
    tidy_model$candidate <- candidate  # Add candidate to the output
    return(tidy_model)
  } else {
    return(NULL)  # Return NULL if no data for the candidate
  }
}

@@

# View the final results
kable(final_results)

@@

## 2. Primary Model (Swing State-Based Model)
This model forecasts electoral outcomes by projecting state-level polling results for each candidate, combining swing state outcomes with votes from solid states.

@@

2. **Predicting Swing State Winners**  
   On election day ($t = 0$), each candidate’s predicted support in state $s$is:
   $$
   P_{H_s}(0) = \alpha_{H_s}, \quad P_{T_s}(0) = \alpha_{T_s}
   $$

   Define $W_s$ as an indicator for the swing state winner:
   $$
   W_s = 
   \begin{cases} 
   1 & \text{if } \alpha_{H_s} > \alpha_{T_s} \\
   0 & \text{otherwise}
   \end{cases}
   $$

3. **Electoral Vote Aggregation**  
   Let $EV_H$ and $EV_T$ represent total electoral votes for Harris and Trump:
   $$
   EV_H = \sum_{s \in S} W_s \cdot EV_s + \sum_{s \in S_H} EV_s
   $$
   $$
   EV_T = \sum_{s \in S} (1 - W_s) \cdot EV_s + \sum_{s \in S_T} EV_s
   $$

4. **Final Election Outcome**  
   The candidate surpassing 270 electoral votes wins:
   $$
   W = 
   \begin{cases} 
   \text{Harris wins} & \text{if } EV_H \geq 270 \\
   \text{Trump wins} & \text{if } EV_T \geq 270 
   \end{cases}
   $$

@@

# Calculate the difference in estimates between candidates
intercept_wide <- intercept_wide %>%
  mutate(difference = `Kamala Harris` - `Donald Trump`)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
solid_state_votes <- data.frame(
  Candidate = c("Harris", "Trump"),
  Solid_State_Votes = c(226, 219)
)
intercept_wide$electoral_votes <- c(6, 11, 10, 15, 19, 16, 16)

@@

total_votes <- solid_state_votes %>%
  left_join(swing_state_results, by = c("Candidate" = "Winner")) %>%
  mutate(Predicted_Swing_State_Votes = ifelse(is.na(Predicted_Swing_State_Votes), 0, Predicted_Swing_State_Votes),
         Total_Predicted_Votes = Solid_State_Votes + Predicted_Swing_State_Votes)

kable(total_votes, caption = "Predicted Electoral Votes for Each Candidate")
```

@@

>>>>

```{r}
data_swing <- data %>% filter(state %in% c("Nevada", "Arizona", "Wisconsin", "Michigan", "Pennsylvania", "North Carolina", "Georgia"))

@@

In this study, we employ two main models to forecast the outcome of the 2024 U.S. Presidential Election using polling data for the candidates Kamala Harris and Donald Trump. The models, a Baseline Model (Winner-Take-All National Model) and a Swing State-Based Model, provide two distinct but complementary approaches to assessing candidate support. The Baseline Model aggregates national polling data to project which candidate is likely to win based on national polling trends and assigns all electoral votes to the candidate with the higher projected national percentage on election day. This approach offers a straightforward prediction of overall voter sentiment across the country and captures the aggregate trend of support for each candidate as the election approaches.

On the other hand, the Swing State-Based Model is designed to provide a more granular projection by focusing on polling data at the state level, particularly for key swing states. This model assumes that the winner in each swing state can be determined by state-specific polling trends, and it projects support for each candidate within these competitive regions. By combining outcomes from both swing states and solid states, this model allows for a more nuanced prediction that aligns with the actual mechanics of the Electoral College. Together, these models are well-suited to address both the general national polling trends and the crucial role of swing states in determining the final electoral outcome.

@@

  if (nrow(candidate_data) > 0) {  # Check if there are rows for the candidate
    model <- lm(pct ~ days_towards_election, data = candidate_data)
    model_summary <- summary(model)
    tidy_model <- data.frame(
      term = rownames(model_summary$coefficients),
      estimate = round(model_summary$coefficients[, "Estimate"], 3),
      std.error = round(model_summary$coefficients[, "Std. Error"], 3),
      t.stat = round(model_summary$coefficients[, "t value"], 3),
      p.value = round(model_summary$coefficients[, "Pr(>|t|)"], 3)
    )
    tidy_model$candidate <- candidate  # Add candidate to the output
    return(tidy_model)
  } else {
    return(NULL)  # Return NULL if no data for the candidate
  }
}

@@

# Combine results for all candidates into a single data frame
final_results_baseline <- bind_rows(results)

# View the final results
kable(final_results_baseline, row.names = FALSE)

@@

## 1. Baseline Model (Winner-Take-All National Model)
The baseline model aggregates national polling data to predict a winner, with all electoral votes awarded to the candidate with the higher national polling percentage.
   Let:
   - $P_H(t)$: Harris’s national polling percentage at $t$ days towards the election.
   - $P_T(t)$: Trump’s national polling percentage at $t$ days towards the election.

@@

Then, our aim is the intercept, $P_H(0) = \alpha_H, \quad P_T(0) = \alpha_T$.

We then draw the prediction result as follows: Harris wins (all electoral votes) if $\alpha_H > \alpha_T$, and Trump wins (all electoral votes) otherwise.

@@

  if (nrow(candidate_data) > 0) {  # Check if there are rows for the candidate
    model <- lm(pct ~ days_towards_election, data = candidate_data)
    model_summary <- summary(model)
    tidy_model <- data.frame(
      term = rownames(model_summary$coefficients),
      estimate = round(model_summary$coefficients[, "Estimate"], 3),
      std.error = round(model_summary$coefficients[, "Std. Error"], 3),
      t.stat = round(model_summary$coefficients[, "t value"], 3),
      p.value = round(model_summary$coefficients[, "Pr(>|t|)"], 3)
    )
    tidy_model$state <- unique(candidate_data$state)  # Add state to the output
    tidy_model$candidate <- candidate  # Add candidate to the output
    return(tidy_model)
  } else {
    return(NULL)  # Return NULL if no data for the candidate
  }
}

@@

# View the final results
kable(final_results, row.names = FALSE)

@@

## 2. Swing State-Based Model
This model forecasts electoral outcomes by projecting state-level polling results for each candidate, combining swing state outcomes with votes from solid states.

@@

Then, we obtain the predicted polling percentage for Harris in state $s:P_{H_s}(0) = \alpha_{H_s},$ and that for Trump in state $s: \quad P_{T_s}(0) = \alpha_{T_s}$. The candidate mentioned above wins in state s if their predicted polling percentage is higher than the other candidate.

2. **Electoral Vote Aggregation**  
  Based on the electoral vote policy in swing states, the winner of the state obtains all the vote from that state. Obtaining the predicted winner of the swing states, we calculate their total electoral votes as two components: their votes from the solid states as mentioned by Electoral Ventures LLC (2024), and the votes from the swing states that would advocate the candidate based on the predictions. 
  Eventually, the candidate surpassing 270 electoral votes in our prediction wins.

@@

# Calculate the difference in estimates between candidates
intercept_wide <- intercept_wide %>%
  mutate(Difference = `Kamala Harris` - `Donald Trump`) %>% 
  rename(State = state)

intercept_wide$`Votes` <- c(6, 11, 10, 15, 19, 16, 16)

@@

```{r}
#| echo: false
#| message: false
#| warning: false
solid_state_votes <- data.frame(
  Candidate = c("Harris", "Trump"),
  Solid_State = c(226, 219)
)

@@

total_votes <- solid_state_votes %>%
  left_join(swing_state_results, by = c("Candidate" = "Winner")) %>%
  mutate(Predicted_Swing_State_Votes = ifelse(is.na(Predicted_Swing_State_Votes), 0, Predicted_Swing_State_Votes),
         total = Solid_State + Predicted_Swing_State_Votes) %>% 
  rename(`Solid State` = Solid_State, `Predicted Swing State` = Predicted_Swing_State_Votes, `Total Predicted Votes` = total)

# View the results
kable(total_votes, caption = "Predicted Electoral Votes for Each Candidate")

@@

Electoral Ventures LLC, (2024). 270toWin - 2024 Presidential Election Interactive Map. 270toWin.com. https://www.270towin.com/
